Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2016-01-07 manuscript-post-acceptance-submission/ensembler.tex
Added manuscript-post-acceptance-submission

\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[mathbf,mathcal]{euler}
%\usepackage{citesort}
\usepackage[caption=false]{subfig}
\usepackage{dcolumn}
\usepackage{boxedminipage}
\usepackage{verbatim}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[group-separator={,}]{siunitx}

@@

\author{Daniel L. Parton}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065}
  %\email{daniel.parton@choderalab.org}
\author{Patrick B. Grinaway}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065}
\author{Sonya M. Hanson}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065}
\author{Kyle A. Beauchamp}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065}
  %\email{patrick.grinaway@choderalab.org}
\author{John D. Chodera}
 \thanks{Corresponding author}
 \email{john.chodera@choderalab.org}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065}

\date{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT/pacs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}

@@

\end{abstract}

@@

Molecular dynamics (MD) simulations have the capability, in principle, to describe the time evolution of a protein in atomistic detail, and have proven themselves to be a useful tool in the study of protein dynamics.
A number of mature software packages and forcefields are now available, and much recent progress has been driven by advances in computing architecture.
For example, many MD packages are now able to exploit GPUs~\cite{eastman:jctc:2012:openmm,walker:jctc:2013:amber-gpu}, which provide greatly improved simulation efficiency per unit cost relative to CPUs, while distributed computing platforms such as Folding@home~\cite{shirts-pande:science:2000:fah}, Copernicus~\cite{pronk:2011:copernicus,pronk:2015:copernicus}, and GPUGrid~\cite{buch:2010:gpugrid}, allow scalability on an unprecedented level.
In parallel, methods for building human-understandable models of protein dynamics from noisy simulation data, such as Markov state modeling (MSM) approaches, are now reaching maturity~\cite{pande-beauchamp-bowman:2010:methods:markov-model-review,noe:jcp:2011:msm-review,chodera-noe:curr-opin-struct-biol:2014:msm-review}.
MSM methods in particular have the advantage of being able to aggregate data from multiple independent MD trajectories, facilitating parallelization of production simulations and thus greatly alleviating overall computational cost.
\color{red}
There also exist a number of mature software packages for comparative modeling of protein structures, in which a target protein sequence is modeled using one or more structures as templates~\cite{moult:proteins:2014:casp10,baker:science:2001:structure-prediction-review,baker:structure:2013:rosettacm,yang:nature-methods:2015:i-tasser-suite}.
\color{black}

\color{red}
However, it remains difficult for researchers to exploit the full variety of available protein sequence data (in simulating groups of related proteins) and structural data (exploiting multiple structures for each protein and its homologs/orthologs) in simulation studies in molecular simulations, largely due to limitations in software architecture.
For example, the preparation of a biomolecular simulation is typically performed manually, encompassing a series of fairly standard (yet time-consuming) steps such as the choice of protein sequence construct and starting structure(s), addition of missing residues and atoms, solvation with explicit water and counterions (and potentially buffer components and cosolvents), choice of simulation parameters (or parameterization schemes for components where parameters do not yet exist),  system relaxation with energy minimization, and one or more short preparatory MD simulations to equilibrate the system and relax the simulation cell.
Due to the laborious and manual nature of this process, simulation studies typically consider only one or a few proteins and starting configurations,  though notable exceptions exist, such as the Dynameomics effort of Daggett and coworkers in which over 100 proteins have been simulated so far using a single initial configuration for each~\cite{daggett:structure:2010:dynameomics}.
Worse still, studies (or collections of studies) that \emph{do} consider multiple proteins often suffer from the lack of consistent best practices in this preparation process, making comparisons between related proteins unnecessarily difficult.

The ability to fully exploit the large quantity of available protein sequence and structural data in biomolecular simulation studies could open up many interesting avenues for research, enabling the study of entire protein families or superfamilies within a single organism or across multiple organisms.
The similarity between members of a given protein family could be exploited to generate arrays of conformational models for related sequences, which could be used as starting configurations to aid sampling in MD simulations.
The conformations captured in structures of related members has been shown to provide useful information about the conformations accessible to all members of the family~\cite{friedland:plos-comput-biol:2009:conformational-dynamics,sali:2012:allostery-modeling}, though energetic differences between individuals will modify the populations and dynamics of individual conformational states.
This approach would be highly beneficial for many MD methods, such as MSM construction, which require global coverage of the conformational landscape to realize their full potential, and would also be particularly useful in cases where structural data is present for only a subset of the members of a protein family.
It would also aid in studying protein families known to have multiple metastable conformations---such as kinases---for which the combined body of structural data for the family may cover a large range of these conformations, while the available structures for any individual member might encompass only one or two distinct conformations.
\color{black}

@@

\color{red}
We anticipate that {\bf Ensembler} will prove to be useful in a number of other ways.
For example, the generated models could represent valuable data sets even without subsequent production simulation, allowing exploration of the conformational diversity present within the available structural data for a given protein family.
Furthermore, automation of simulation preparation provides an excellent opportunity to make concrete certain "best practices", such as the choice of simulation parameters, approach to the treatment of protonation states, treatment of cofactors and structural ions, and pre-simulation refinement and equilibration procedures.
While the current version of {\bf Ensembler} only codifies some of these choices as default parameters, its modular nature allows additional stages to be easily added in the future.
\color{black}

@@

\begin{figure*}[tb]
    \includegraphics[width=1.0\textwidth]{pipeline/pipeline2}

  \caption{{\bf Diagrammatic representation of the stages of the Ensembler pipeline and illustrative statistics for modeling all human tyrosine kinase catalytic domains.}
  On the left, the various stages of the {\bf Ensembler} pipeline are shown.
  The red labels indicate the corresponding text description provided for each stage in the Design and Implementation section.
  On the right, the number of viable models surviving each stage of the pipeline is shown for the 93 target TK domains and for two representative individual TK domains (\emph{SRC} and \emph{ABL}).
  Typical timings on a computer cluster (containing Intel Xeon E5-2665 2.4GHz hyperthreaded processors and NVIDIA GTX-680 or GTX-Titan GPUs) is reported to illustrate resource requirements per model for modeling the entire set of tyrosine kinases.
  Note that \emph{CPU-h} denotes the number of hours consumed by the equivalent of a single CPU hyperthread and \emph{GPU-h} on a single GPU---parallel execution via MPI reduces wall clock time nearly linearly.
  }
  \label{figure:pipeline}
\end{figure*}

\subsection{Target selection and retrieval}

@@

\subsection{Template selection and retrieval}

{\bf Ensembler} uses comparative modeling to build models, and as such requires a set of structures to be used as templates.
The second stage thus entails the selection of templates and storage of associated sequences, structures, and identifiers.
These templates can be specified manually, or using the {\tt ensembler gather\_templates} subcommand to automatically select templates based on a search of the Protein Data Bank (PDB) or UniProt.
\color{red}
A recommended approach is to select templates from UniProt which belong to the same protein family as the targets, establishing homology and some degree of sequence identity between targets and templates.
\color{black}

The {\tt ensembler gather\_templates} subcommand provides methods for selecting template structures from either UniProt or the PDB (\url{http://www.rcsb.org/pdb}), specified by the {\tt -{}-gather\_from} flag.
\color{red}
Both methods select templates at the level of PDB chains---a PDB structure containing multiple chains with identical sequence spans (e.g.~for crystals with non-crystallographic symmetry giving rise to independent conformations of the protein within the asymmetric unit) would thus give rise to multiple template structures.
\color{black}

@@

\subsection{Template refinement}

@@

\subsection{Alignment and comparative modeling}

@@

While MODELLER's automodel function can generate its own alignments automatically, a standalone function was preferable for reasons of programming convenience.
\color{red}
As such, we implemented pairwise alignment functionality using the BioPython {\tt pairwise2} module~\cite{cock:2009:biopython}---which uses a dynamic programming algorithm---with the PAM 250 scoring matrix of Gonnet \textit{et al.}~\cite{gonnet:science:1992:exhaustive-matching}, though other choices of scoring matrices available within the module can be selected.
\color{black}
The alignments are carried out with the {\tt align} subcommand, prior to the modeling step which is carried out with the {\tt build\_models} subcommand.
The {\tt align} subcommand also writes a list of the sequence identities for each template to a text file, and this can be used to select models from a desired range of sequence identities.
The {\tt build\_models} subcommand and all subsequent pipeline functions have a {\tt -{}-template\_seqid\_cutoff} flag which can be used to select only models with sequence identities greater than the given value.
We also note that alternative approaches could be used for the alignment stage.
For example, multiple sequence alignment algorithms~\cite{thompson:2011:msa-review}, allow alignments to be guided using sequence data from across the entire protein family of interest, while (multiple) structural alignment algorithms such as MODELLER's {\tt salign} routine~\cite{fiser:prot-sci:2000:modeller,sali:jmb:1993:modeller}, PROMALS3D~\cite{pei:2008:promals3d}, and Expresso and 3DCoffee~\cite{expresso,3dcoffee}, can additionally exploit structural data.
{\bf Ensembler's} modular architecture facilitates the implementation of alternative alignment approaches, and we plan to implement some of these in future versions, to allow exploration of the influence of different alignment methods on model quality.

@@

\color{red}
At this time, the alignment and modeling functions cannot be used to model non-standard amino acids, though we plan to be able to provide this functionality in future versions.
Note that the Ensembler functions for target and template selection only include standard amino acids which match the UniProt canonical isoform sequence, and thus any set of targets and templates selected this way should be compatible with the Ensembler alignment and modeling functions.
\color{black}

@@

\color{red}
\subsection{Refinement of models and filtering of poor models by simulation}

A number of refinement methods have been developed to help guide comparative modeling techniques toward more "native-like" and physically consistent conformations~\cite{maccallum:prot:2011:casp-refinement,zhang:curr-opin-struct-biol:2009:structure-prediction}.
Both short~\cite{zhang:curr-opin-struct-biol:2009:structure-prediction} and long~\cite{raval:prot:2012:long-timescale-md-refinement} molecular dynamics simulations have been employed for this purpose.
Here, we utilize short molecular dynamic simulations for two purposes: both to slightly relax the initial comparative models and to eliminate those comparative models that result in highly implausible conformations.
This is especially critical here due to the inclusion of even very low sequence identity template structures.
We stress that the limited refinement by molecular simulation here is primarily intended as initial relaxation and filtering stages, where implausible models might cause simulations to immediately fail, crash, or generate implausibly high energies or unstable dynamics.
Exploration of conformational dynamics to derive MSMs, for example, will inevitably require orders of magnitude more simulation effort---very likely tens of microseconds to milliseconds of aggregate dynamics~\cite{pande-beauchamp-bowman:2010:methods:markov-model-review,chodera-noe:curr-opin-struct-biol:2014:msm-review}.
\color{black}

@@

\color{red}
For some studies, it may be useful to specify the protonation states of individual amino acids, rather than rely only on automatic protonation state assignment by OpenMM.
The user can do this by listing the residue numbers and their protonation states in a configuration file ({\tt manual\_overrides.yaml}).
The necessary formatting for the configuration file is specified in the software documentation, and a template file is written when initializing an {\bf Ensembler} project.
Protonation states are specified by naming the appropriate residue variant type in the force field, e.g. {\tt `ASH'} for an aspartic acid residue, as opposed to the aspartate base {\tt `ASP'}.
Any residues which do not have specific protonation states listed in the configuration file will have protonation states assigned automatically by OpenMM.
\color{red}
Note that {\bf Ensembler} currently only supports residue definitions provided by the forcefield definition files---it does not yet have the ability to derive new forcefield parameters for uncommon amino acids, cofactors, or ions not provided by the forcefield.
\color{black}

\subsection{Solvation and NPT equilibration}

@@

\subsection*{{\color{red} Model validation with MolProbity score}}

\color{red}
{\bf Ensembler} provides a function for validating model quality and filtering models using MolProbity~\cite{molprobity:2007,molprobity:2010}---a widely used tool for validation of protein models, which provides a numerical score derived from features such as steric clashes between atoms, bond geometry, Ramachandran angles, sidechain rotamer outliers, backbone deviations, and the presence of cis-peptides.
This function is accessed via the {\tt validate} subcommand, which for a given target will output a text file containing a list of model IDs sorted by validation score.
The optional {\tt -{}-modeling\_stage} flag specifies which of the three main Ensembler modeling stages to validate---the initial comparative modeling stage, the implicit MD refinement stage, or the explicit MD refinement stage.
If this flag is not used, Ensembler defaults to selecting the latest stage for which models have been generated.
The output text file can be used to filter models based on validation score, for example by using the {\tt package\_models} subcommand.
Protein model validation is a challenging problem and an active area of research for many groups, including the developers of {\bf Ensembler}.
We plan to implement further validation methods in future versions of Ensembler.

\color{black}

@@

\color{red}
{\bf Ensembler} provides a packaging module which can be used to prepare models for subsequent downstream use, such as the use of distributed or cluster computing resources for the generation of MSMs~\cite{pande-beauchamp-bowman:2010:methods:markov-model-review,noe:jcp:2011:msm-review,chodera-noe:curr-opin-struct-biol:2014:msm-review}.
The {\tt package\_models} subcommand currently provides functions (specified via the {\tt -{}-package\_for} flag) for compressing models in preparation for data transfer, or for organizing them with the appropriate directory and file structure for production simulation on the distributed computing platform Folding@home~\cite{shirts-pande:science:2000:fah}.
The module could easily be extended to add methods for preparing models for other purposes.
For example, production simulations could alternatively be run using Copernicus~\cite{pronk:2011:copernicus,pronk:2015:copernicus}---a framework for performing parallel adaptive MD simulations--- or GPUGrid~\cite{buch:2010:gpugrid}---a distributing computing platform which relies on computational power voluntarily donated by the owners of nondedicated GPU-equipped computers.

\color{red}
An important use of the packaging stage is to filter models based on model quality.
At the current time, the available filtering options are based on either target-template sequence identity or MolProbity validation score.
The {\tt package\_models} subcommand includes optional flags for specifying a sequence identity cutoff (so that only models with a target-template sequence identity above the specified percentage are chosen), a MolProbity validation score cutoff (to choose only models with lower validation scores, which indicate better model quality), or a MolProbity validation score percentile (to choose only models with validation scores lower than the value at the given percentile).

@@

\color{red}
We stress that, despite evidence suggesting that there is a correspondence between solution-state dynamics and structural diversity of related template proteins~\cite{friedland:plos-comput-biol:2009:conformational-dynamics}, all models---especially those derived from low sequence identity templates---are not necessarily representative of conformations thermally accessible to the template proteins of interest.
Care must be exercised in the use and analysis of these models.
\color{black}

@@

\color{red}
Figure~\ref{figure:nstructures-per-tk-target} shows the number of PDB structures available for each of the 93 target TK domains.
While a number of experimental structures are available for some TK domains, many TKs have few or no structures. 
{\bf Ensembler} thus helps to overcome this unequal distribution of structural information when building protein models for simulation by exploiting homologous structural data from a wider range of protein kinase domains and species.
\color{black}

\begin{figure*}[tbp]
    \includegraphics[width=\textwidth]{nstructures_per_tk_target/nstructures_per_tk_target.pdf}
    \caption{{\bf Number of PDB structures available for each TK target.}
    \color{red}
    Data is shown for each of the 93 TK kinase domains, sorted in order of the number of available PDB structures for each domain. 
    The labels indicate the UniProt name for the target protein plus an index for the kinase domain (three of the selected proteins have two kinase domains). 
    Each PDB chain is counted individually, and only chains which contain the target domain are counted.
}
    \label{figure:nstructures-per-tk-target}
\end{figure*}

@@

\begin{figure}[tbp]
    \includegraphics[width=\textwidth]{loopmodel_analysis/nmissing_resis_distributions.pdf}
    \caption{{\bf Distributions for the number of missing residues in the TK templates.}
    \color{red}
    \emph{Upper:} The number of missing residues per template, for all templates (blue) and for only those templates for which template remodeling with the {\tt loopmodel} subcommand failed (red).
    Templates for which remodeling failed had a median of 20 missing residues, compared to a median of 14 missing residues for templates for which remodeling was successful.
    \emph{Lower:} The number of residues in each missing loop, for all templates.    
}
    \label{figure:loopmodel-nmissing-residues}
\end{figure}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[mathbf,mathcal]{euler}
%\usepackage{citesort}
\usepackage[caption=false]{subfig}
\usepackage{dcolumn}
\usepackage{boxedminipage}
\usepackage{verbatim}
%\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[draft]{hyperref}
\usepackage[group-separator={,}]{siunitx}

@@

\author{Daniel L. Parton}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065 \\
Corresponding author: john.chodera@choderalab.org
}
  %\email{daniel.parton@choderalab.org}
\author{Patrick B. Grinaway}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065 \\
Corresponding author: john.chodera@choderalab.org
}
\author{Sonya M. Hanson}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065 \\
Corresponding author: john.chodera@choderalab.org
}
\author{Kyle A. Beauchamp}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065 \\
Corresponding author: john.chodera@choderalab.org
}
  %\email{patrick.grinaway@choderalab.org}
\author{John D. Chodera}
 %\thanks{Corresponding author}
 %\email{john.chodera@choderalab.org}
  \affiliation{Computational Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York, NY 10065 \\
Corresponding author: john.chodera@choderalab.org
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT/pacs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{abstract}

\section*{abstract}

@@

% \end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AUTHOR SUMMARY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Author Summary}
\label{section:author_summary}

Proteins are the workhorses of the human body, and are involved in essentially every biological process.
Many diseases are caused by proteins malfunctioning.
To understand how a protein functions, it is necessary to know its physical properties.
The field of structural biology provides many techniques for determining the three-dimensional structure of a protein.
The dynamics of a protein, i.e. the way it moves, are of equal importance, but are more difficult to uncover with traditional experimental techniques.
Computer simulations are an effective alternative method for understanding protein dynamics, but require experimental structural information as a starting point.
While recent advances in genomics and experimental techniques have provided a wealth of such structural data, the appropriate software for using this data effectively has been lacking.
To tackle this problem, we have developed a software package called {\bf Ensembler}, which allows a user to automatically select appropriate experimentally derived structures for a given protein or family of proteins, and to use them to prepare a series of simulations.
The resultant simulation data can then used to investigate the dynamics of the protein(s) in question, and their involvement in disease.

@@

Molecular dynamics (MD) simulations have the capability, in principle, to describe the time evolution of a protein in atomistic detail, and have proven themselves to be a useful tool in the study of protein dynamics.
A number of mature software packages and forcefields are now available, and much recent progress has been driven by advances in computing architecture.
For example, many MD packages are now able to exploit GPUs~\cite{eastman:jctc:2012:openmm,walker:jctc:2013:amber-gpu}, which provide greatly improved simulation efficiency per unit cost relative to CPUs, while distributed computing platforms such as Folding@home~\cite{shirts-pande:science:2000:fah}, Copernicus~\cite{pronk:2011:copernicus,pronk:2015:copernicus}, and GPUGrid~\cite{buch:2010:gpugrid}, allow scalability on an unprecedented level.
In parallel, methods for building human-understandable models of protein dynamics from noisy simulation data, such as Markov state modeling (MSM) approaches, are now reaching maturity~\cite{pande-beauchamp-bowman:2010:methods:markov-model-review,noe:jcp:2011:msm-review,chodera-noe:curr-opin-struct-biol:2014:msm-review}.
MSM methods in particular have the advantage of being able to aggregate data from multiple independent MD trajectories, facilitating parallelization of production simulations and thus greatly alleviating overall computational cost.
There also exist a number of mature software packages for comparative modeling of protein structures, in which a target protein sequence is modeled using one or more structures as templates~\cite{moult:proteins:2014:casp10,baker:science:2001:structure-prediction-review,baker:structure:2013:rosettacm,yang:nature-methods:2015:i-tasser-suite}.

However, it remains difficult for researchers to exploit the full variety of available protein sequence data (in simulating groups of related proteins) and structural data (exploiting multiple structures for each protein and its homologs/orthologs) in simulation studies in molecular simulations, largely due to limitations in software architecture.
For example, the preparation of a biomolecular simulation is typically performed manually, encompassing a series of fairly standard (yet time-consuming) steps such as the choice of protein sequence construct and starting structure(s), addition of missing residues and atoms, solvation with explicit water and counterions (and potentially buffer components and cosolvents), choice of simulation parameters (or parameterization schemes for components where parameters do not yet exist),  system relaxation with energy minimization, and one or more short preparatory MD simulations to equilibrate the system and relax the simulation cell.
Due to the laborious and manual nature of this process, simulation studies typically consider only one or a few proteins and starting configurations,  though notable exceptions exist, such as the Dynameomics effort of Daggett and coworkers in which over 100 proteins have been simulated so far using a single initial configuration for each~\cite{daggett:structure:2010:dynameomics}.
Worse still, studies (or collections of studies) that \emph{do} consider multiple proteins often suffer from the lack of consistent best practices in this preparation process, making comparisons between related proteins unnecessarily difficult.

The ability to fully exploit the large quantity of available protein sequence and structural data in biomolecular simulation studies could open up many interesting avenues for research, enabling the study of entire protein families or superfamilies within a single organism or across multiple organisms.
The similarity between members of a given protein family could be exploited to generate arrays of conformational models for related sequences, which could be used as starting configurations to aid sampling in MD simulations.
The conformations captured in structures of related members has been shown to provide useful information about the conformations accessible to all members of the family~\cite{friedland:plos-comput-biol:2009:conformational-dynamics,sali:2012:allostery-modeling}, though energetic differences between individuals will modify the populations and dynamics of individual conformational states.
This approach would be highly beneficial for many MD methods, such as MSM construction, which require global coverage of the conformational landscape to realize their full potential, and would also be particularly useful in cases where structural data is present for only a subset of the members of a protein family.
It would also aid in studying protein families known to have multiple metastable conformations---such as kinases---for which the combined body of structural data for the family may cover a large range of these conformations, while the available structures for any individual member might encompass only one or two distinct conformations.

@@

We anticipate that {\bf Ensembler} will prove to be useful in a number of other ways.
For example, the generated models could represent valuable data sets even without subsequent production simulation, allowing exploration of the conformational diversity present within the available structural data for a given protein family.
Furthermore, automation of simulation preparation provides an excellent opportunity to make concrete certain "best practices", such as the choice of simulation parameters, approach to the treatment of protonation states, treatment of cofactors and structural ions, and pre-simulation refinement and equilibration procedures.
While the current version of {\bf Ensembler} only codifies some of these choices as default parameters, its modular nature allows additional stages to be easily added in the future.

@@

\begin{figure*}[tb]
    %\includegraphics[width=1.0\textwidth]{pipeline/pipeline2}

  \caption{{\bf Diagrammatic representation of the stages of the Ensembler pipeline and illustrative statistics for modeling all human tyrosine kinase catalytic domains.}
  On the left, the various stages of the {\bf Ensembler} pipeline are shown.
  The red labels indicate the corresponding text description provided for each stage in the Design and Implementation section.
  On the right, the number of viable models surviving each stage of the pipeline is shown for the 93 target TK domains and for two representative individual TK domains (\emph{SRC} and \emph{ABL}).
  Typical timings on a computer cluster (containing Intel Xeon E5-2665 2.4GHz hyperthreaded processors and NVIDIA GTX-680 or GTX-Titan GPUs) is reported to illustrate resource requirements per model for modeling the entire set of tyrosine kinases.
  Note that \emph{CPU-h} denotes the number of hours consumed by the equivalent of a single CPU hyperthread and \emph{GPU-h} on a single GPU---parallel execution via MPI reduces wall clock time nearly linearly.
  \label{figure:pipeline}
  }
\end{figure*}

\subsection{Target selection and retrieval (A)}

@@

\subsection{Template selection and retrieval (B)}

{\bf Ensembler} uses comparative modeling to build models, and as such requires a set of structures to be used as templates.
The second stage thus entails the selection of templates and storage of associated sequences, structures, and identifiers.
These templates can be specified manually, or using the {\tt ensembler gather\_templates} subcommand to automatically select templates based on a search of the Protein Data Bank (PDB) or UniProt.
A recommended approach is to select templates from UniProt which belong to the same protein family as the targets, establishing homology and some degree of sequence identity between targets and templates.

The {\tt ensembler gather\_templates} subcommand provides methods for selecting template structures from either UniProt or the PDB (\url{http://www.rcsb.org/pdb}), specified by the {\tt -{}-gather\_from} flag.
Both methods select templates at the level of PDB chains---a PDB structure containing multiple chains with identical sequence spans (e.g.~for crystals with non-crystallographic symmetry giving rise to independent conformations of the protein within the asymmetric unit) would thus give rise to multiple template structures.

@@

\subsection{Template refinement (C)}

@@

\subsection{Alignment and comparative modeling (D)}

@@

While MODELLER's automodel function can generate its own alignments automatically, a standalone function was preferable for reasons of programming convenience.
As such, we implemented pairwise alignment functionality using the BioPython {\tt pairwise2} module~\cite{cock:2009:biopython}---which uses a dynamic programming algorithm---with the PAM 250 scoring matrix of Gonnet \textit{et al.}~\cite{gonnet:science:1992:exhaustive-matching}, though other choices of scoring matrices available within the module can be selected.
The alignments are carried out with the {\tt align} subcommand, prior to the modeling step which is carried out with the {\tt build\_models} subcommand.
The {\tt align} subcommand also writes a list of the sequence identities for each template to a text file, and this can be used to select models from a desired range of sequence identities.
The {\tt build\_models} subcommand and all subsequent pipeline functions have a {\tt -{}-template\_seqid\_cutoff} flag which can be used to select only models with sequence identities greater than the given value.
We also note that alternative approaches could be used for the alignment stage.
For example, multiple sequence alignment algorithms~\cite{thompson:2011:msa-review}, allow alignments to be guided using sequence data from across the entire protein family of interest, while (multiple) structural alignment algorithms such as MODELLER's {\tt salign} routine~\cite{fiser:prot-sci:2000:modeller,sali:jmb:1993:modeller}, PROMALS3D~\cite{pei:2008:promals3d}, and Expresso and 3DCoffee~\cite{expresso,3dcoffee}, can additionally exploit structural data.
{\bf Ensembler's} modular architecture facilitates the implementation of alternative alignment approaches, and we plan to implement some of these in future versions, to allow exploration of the influence of different alignment methods on model quality.

@@

At this time, the alignment and modeling functions cannot be used to model non-standard amino acids, though we plan to be able to provide this functionality in future versions.
Note that the Ensembler functions for target and template selection only include standard amino acids which match the UniProt canonical isoform sequence, and thus any set of targets and templates selected this way should be compatible with the Ensembler alignment and modeling functions.

@@

\subsection{Refinement of models and filtering of poor models by simulation (E)}

A number of refinement methods have been developed to help guide comparative modeling techniques toward more "native-like" and physically consistent conformations~\cite{maccallum:prot:2011:casp-refinement,zhang:curr-opin-struct-biol:2009:structure-prediction}.
Both short~\cite{zhang:curr-opin-struct-biol:2009:structure-prediction} and long~\cite{raval:prot:2012:long-timescale-md-refinement} molecular dynamics simulations have been employed for this purpose.
Here, we utilize short molecular dynamic simulations for two purposes: both to slightly relax the initial comparative models and to eliminate those comparative models that result in highly implausible conformations.
This is especially critical here due to the inclusion of even very low sequence identity template structures.
We stress that the limited refinement by molecular simulation here is primarily intended as initial relaxation and filtering stages, where implausible models might cause simulations to immediately fail, crash, or generate implausibly high energies or unstable dynamics.
Exploration of conformational dynamics to derive MSMs, for example, will inevitably require orders of magnitude more simulation effort---very likely tens of microseconds to milliseconds of aggregate dynamics~\cite{pande-beauchamp-bowman:2010:methods:markov-model-review,chodera-noe:curr-opin-struct-biol:2014:msm-review}.

@@

For some studies, it may be useful to specify the protonation states of individual amino acids, rather than rely only on automatic protonation state assignment by OpenMM.
The user can do this by listing the residue numbers and their protonation states in a configuration file ({\tt manual\_overrides.yaml}).
The necessary formatting for the configuration file is specified in the software documentation, and a template file is written when initializing an {\bf Ensembler} project.
Protonation states are specified by naming the appropriate residue variant type in the force field, e.g. {\tt `ASH'} for an aspartic acid residue, as opposed to the aspartate base {\tt `ASP'}.
Any residues which do not have specific protonation states listed in the configuration file will have protonation states assigned automatically by OpenMM.
Note that {\bf Ensembler} currently only supports residue definitions provided by the forcefield definition files---it does not yet have the ability to derive new forcefield parameters for uncommon amino acids, cofactors, or ions not provided by the forcefield.

\subsection{Solvation and NPT equilibration (F)}

@@

\subsection*{{Model validation with MolProbity score}}

{\bf Ensembler} provides a function for validating model quality and filtering models using MolProbity~\cite{molprobity:2007,molprobity:2010}---a widely used tool for validation of protein models, which provides a numerical score derived from features such as steric clashes between atoms, bond geometry, Ramachandran angles, sidechain rotamer outliers, backbone deviations, and the presence of cis-peptides.
This function is accessed via the {\tt validate} subcommand, which for a given target will output a text file containing a list of model IDs sorted by validation score.
The optional {\tt -{}-modeling\_stage} flag specifies which of the three main Ensembler modeling stages to validate---the initial comparative modeling stage, the implicit MD refinement stage, or the explicit MD refinement stage.
If this flag is not used, Ensembler defaults to selecting the latest stage for which models have been generated.
The output text file can be used to filter models based on validation score, for example by using the {\tt package\_models} subcommand.
Protein model validation is a challenging problem and an active area of research for many groups, including the developers of {\bf Ensembler}.
We plan to implement further validation methods in future versions of Ensembler.

@@

{\bf Ensembler} provides a packaging module which can be used to prepare models for subsequent downstream use, such as the use of distributed or cluster computing resources for the generation of MSMs~\cite{pande-beauchamp-bowman:2010:methods:markov-model-review,noe:jcp:2011:msm-review,chodera-noe:curr-opin-struct-biol:2014:msm-review}.
The {\tt package\_models} subcommand currently provides functions (specified via the {\tt -{}-package\_for} flag) for compressing models in preparation for data transfer, or for organizing them with the appropriate directory and file structure for production simulation on the distributed computing platform Folding@home~\cite{shirts-pande:science:2000:fah}.
The module could easily be extended to add methods for preparing models for other purposes.
For example, production simulations could alternatively be run using Copernicus~\cite{pronk:2011:copernicus,pronk:2015:copernicus}---a framework for performing parallel adaptive MD simulations--- or GPUGrid~\cite{buch:2010:gpugrid}---a distributing computing platform which relies on computational power voluntarily donated by the owners of nondedicated GPU-equipped computers.

An important use of the packaging stage is to filter models based on model quality.
At the current time, the available filtering options are based on either target-template sequence identity or MolProbity validation score.
The {\tt package\_models} subcommand includes optional flags for specifying a sequence identity cutoff (so that only models with a target-template sequence identity above the specified percentage are chosen), a MolProbity validation score cutoff (to choose only models with lower validation scores, which indicate better model quality), or a MolProbity validation score percentile (to choose only models with validation scores lower than the value at the given percentile).

@@

We stress that, despite evidence suggesting that there is a correspondence between solution-state dynamics and structural diversity of related template proteins~\cite{friedland:plos-comput-biol:2009:conformational-dynamics}, all models---especially those derived from low sequence identity templates---are not necessarily representative of conformations thermally accessible to the template proteins of interest.
Care must be exercised in the use and analysis of these models.

@@

Figure~\ref{figure:nstructures-per-tk-target} shows the number of PDB structures available for each of the 93 target TK domains.
While a number of experimental structures are available for some TK domains, many TKs have few or no structures. 
{\bf Ensembler} thus helps to overcome this unequal distribution of structural information when building protein models for simulation by exploiting homologous structural data from a wider range of protein kinase domains and species.

\begin{figure*}[tbp]
    %\includegraphics[width=\textwidth]{nstructures_per_tk_target/nstructures_per_tk_target.pdf}
    \caption{{\bf Number of PDB structures available for each TK target.}
    Data is shown for each of the 93 TK kinase domains, sorted in order of the number of available PDB structures for each domain. 
    The labels indicate the UniProt name for the target protein plus an index for the kinase domain (three of the selected proteins have two kinase domains). 
    Each PDB chain is counted individually, and only chains which contain the target domain are counted.
    \label{figure:nstructures-per-tk-target}
}
\end{figure*}

@@

\begin{figure}[tbp]
    %\includegraphics[width=\textwidth]{loopmodel_analysis/nmissing_resis_distributions.pdf}
    \caption{{\bf Distributions for the number of missing residues in the TK templates.}
    \emph{Upper:} The number of missing residues per template, for all templates (blue) and for only those templates for which template remodeling with the {\tt loopmodel} subcommand failed (red).
    Templates for which remodeling failed had a median of 20 missing residues, compared to a median of 14 missing residues for templates for which remodeling was successful.
    \emph{Lower:} The number of residues in each missing loop, for all templates.    
    \label{figure:loopmodel-nmissing-residues}
}
\end{figure}
